GEEPERs: Principal Stratification using Principal Scores and Stacked Estimating Equations

### 2022-09-30 writeUps/paper.tex
update paper

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
\usepackage{amsmath,amssymb}
\usepackage{bm}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{fullpage}
\usepackage{makecell}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{../../../generalCode/numdef}

@@

Randomized experiments easily admit subgroup analysis---estimation of treatment effects for a subset of the study population---when those subgroups are defined at baseline, and not themselves a function of treatment assignment.
A simple, popular approach to estimating subgroup effects is to use ordinary least squares (OLS) regression, regressing the outcome on a treatment indicator interacted with an indicator for subgroup membership.
However, when subgroup membership is a function of a post-treatment variable, then it may itself be affected by the treatment.
Contrats between treatment and control conditions within those subgroups may not have a causal interpretation at all---and when they do, the same techniques that work for subgroups defined at baseline (such as OLS with interactions) will often produce inconsistant estimates.
Principal stratification \citep{frangakis} is a broad framework for defining subgroup effects for post-treatment subgroups, based on the notition of \emph{potential} subgroups called principal strata, that is, groups of subjects defined based on the values a post-treatment variable would potentially take if the subject is assigned to one treatment condition or another.
Since principal strata membership is based on subjects' responses to both factual and counter-factual treatment assignemtns, it is, in general unobserved.
Nevertheless, researchers have devised estimators of principal effects, as average effects within principal strata are called, to address a broad range of applied statistical problems, including non-compliance in randomized experiments, the evaluation of surrogate outcomes \citep{}, the investigation of causal mechanisms \citep{lidsayPage}, and sample attrition \citep{zhouRubin, pengDing2011}.

Because principal strata membership is generally unobserved, principal effects are not always identified, and when they are, estimation can still be difficult.
Typically, principal effects are identified and estimated based on strong structural assumptions \citep[e.g.][]{air} and/or parametric modeling \citep[e.g.][]{imbensRubin98}.
Parametric models for principal effects are often hard to specify and hard or impossible to test.
Even when they are well-specified they can yield estimates that are severely biased and misleading \citep{griffin,feller}.
Another route to identification and estimation of principal effects leads through auxilliary data, including baseline covariates \citep{} or secondary outcomes \citep{mealli}.
Of particular importance is the ``principal score,'' \citep{jo,ding,feller}, or the probability that a subject belongs to a particular principal stratum, as a function of pre-treatment covariates.
Under a strong, untestable assumption called ``principal ignorability,'' which is akin to the no-confounding assumption underlying observational studies, principal scores can be tranformed into weights, which in turn form an unbiased estimate of principal effects.

@@

This paper focuses on a special case of principal stratification---one-way, binary non-compliance---that is, there are only two principal strata, and principal stratum membership is observable in one treatment arm, but not in the other.
This scenario would occur if, say, subjects in the control arm of a study had no access to the treatment, and not all subjects assigned to the treatment arm actually received the full treatment.
While such a setup may represent the simpliest and easiest principal stratification problem, we believe it is a common scenario.
Moreover, we believe that the methods developed here can extend to more complex scenarios, which we leave to future work.

@@

\section{Background}
In a randomized experiment with two conditions, let $Z_i=0$ if subject $i$, $i=1,\dots,n$ is assigned to the treatment condition, and let $Z_i=0$ otherwise.
If interest is in the effect of $Z$ on an outcome $Y$, under the ``stable unit treatment value assumption'' \citep{sutva} of no interference between units and no hidden versions of the treatment, define potential outcomes \citep{neyman,rubin} $\yti$ as the value of $Y$ that $i$ would exhibit if $Z_i=1$, and let $\yci$ be the value if $Z_i=0$. Then the observed value of $Y$ satisfies $Y_i=Z_i\yti+(1-Z_i)\yci$, and the treatment effect of $Z$ on $Y$ for stubject $i$ may be defined as $\tau_i=\yti-\yci$. The average treatment effect (ATE) is $\EE[\tau]$. %

Let $S_i\in\{0,1\}$ be a measurement on subject $i$ taken subsequent to treatment assignment, so that $S$ may be affected by $Z$.
Then $S$ itself has potential values $\sti$ and $\sci$, which $i$ would exibit if $Z_i=1$ or $Z_i=0$, respectively.
Assume ``strong monotonicity'' \citep[c.f.][]{dingLu} (also referred to as ``one-way noncompliance'' \citealt{imbensRubin}):
\begin{ass}[Strong Monotonicity]\label{ass:SM}
\begin{equation*}%\label{eq:strongMonotonicity}
  \sci=0\text{ for all }i
\end{equation*}
\end{ass}
Like potential outcomes $\yti$ and $\yci$, $\sci$ and $\sti$ are defined for all subjects, though $\sci$ is only observed when $Z_i=0$ and $\sti$ is only observed when $Z_i=1$. Under strong monotonicity, since $S$ is binary, every subject belongs to one of two ``principal strata,'' $\sti=1$ or $\sti=0$.
The average treatment effect in each principal stratum, $\eff1=\EE[\tau|\st=1]$ or $\eff1=\EE[\tau|\st=0]$, is called a principal effect. %
The challange in estimating principal effects is due to the fact that strata membership $\sti$ is only observed when $Z_i=1$, in which case $\sti=S_i$, the observed value of $S$.

@@

Under Assumptions \ref{ass:sm} and \ref{ass:rand}, conditional means $\mut0$ and $\mut1$ are non-parametrically identified.
In particular, the means of observed outcomes for subjects with $Z=1$ and $S=0$ or $S=1$ are unbiased for $\mut0$ and $\mut1$, respectively, since when $Z_i=1$, $Y_i=\yti$ and $S_i=\sti$.
In contrast, $\muc0$ and $\muc1$ are not fully identified without further assumptions because $\sti$ and $\yci$ are never observed simultaneously (though they may be nonparametrically bounded; see, e.g., \citealt{bounding}).
When $S$ is a measure of compliance, and $\tau_i=0$ whenever $\sti=0$ (the ``exclusion restriction''), $\muc1$---and hence $\eff1$---may be estimated with instrumental variables methods \citep{angrist1996identification}.
In this case, $\eff1$ is often called the ``local average treatment effect'' or the average effect of the treatment on the treated.
We will briefly review two approaches to estimating $\muc0$ and $\muc1$ that do not assume the exclusion restriction---finite mixture modeling and weighting---with some comments on the role for covariates in-between.

@@

In order for principal scores to be useful, they need to vary with $\bm{x}$. In particular, we will assume the following:
\begin{ass}[Variable Principal Scores]\label{ass:vps}
  \begin{equation*}
    Var_{\bx} (\pp)>0
  \end{equation*}
\end{ass}

@@

Alternatively, assume ``principal ignorability'' \citep{jo2009use,dingLu}:
\begin{ass}[Principal Ignorability]\label{ass:PI}
\begin{equation}\label{eq:pi}
  Y_{C}\independent \st |\bxs
\end{equation}
\end{ass}
or a somewhat weaker version, $\EE[Y_C|\st,\bxs]=\EE[Y_C|\bxs]$ \citep{feller2017principal}.
That is, the principal stratum is unrelated to control potential outcomes conditional on covariates.
Principal ignorability \eqref{eq:pi} is reminiscent of ignorability assumptions typical in observational studies \citep[e.g.][]{rosenbaum2002observational}.
Both require a rich set of covariates and background knowledge of the determinants of $S$; in particular, an unobserved covariate $u$ that is correlated with both $\st$ and $Y_C$ would invalidate \eqref{eq:pi}.

@@

We begin by introducing an assumption that we term ``Covariate Ignorability,'' or CI. It is a slighty weaker version of \citet{jiangDing2021}'s ``auxilliary indpendence'' assumption.
\begin{ass}[Covariate Ignorability]\label{ass:ci}
\begin{equation}\label{eq:assumption}
\EE[\yci|\bxsi,\sti]=\EE[\yci|\sti]=\muc0\text{ or }\muc1
\end{equation}
\end{ass}
i.e. $Y_C$ is mean-independent of $\bxs$ conditional on $\st$.
Under CI, covariates are not informative of the mean of $Y_C$ within principal strata.
As stated above, CI will rarely be plausible.
However, in some circumstances researchers can identify a subset of observed covariates that satisfy CI, and use those covariates in estimation.

It turns out that under strong monotonicity, randomization, and CI, and given a set of principal scores, the two principal effects $\eff0$ and $\eff1$ can be estimted by a simple OLS regression.
To see how, we'll %first
derive the estimating equations for $\muc0$, $\muc1$, $\mut0$, and $\mut1$, and show that they are equivalent to estimating equations for a model fit by OLS.% assuming that principal scores are known exactly, and then generalize to the case in which they are estimated using data from the treatment group.
The details of the calculations and the proofs are in the appendix.

The argument stems from a set of expressions for expectations, summarized in the following lemma:
\begin{lemma}\label{lemma:expectations}
  Under Assumptions \ref{ass:sm}, \ref{ass:rand}, \ref{ass:vps}, and \ref{ass:ci}, %if principal scores $\ppi$, for $i:Z_i=0$ are known exactly, then:
  \begin{equation*}
    \begin{split}
      \EE[Y|Z=0]=&\muc0+\EE\pp(\muc1-\muc0)\\
      \EE[Y\pp|Z=0]&=\EE[\pp]\muc0+\EE[\pp^2](\muc1-\muc0)\\
      \EE[Y|Z=1]&=\mut0+\EE[\st](\mut1-\mut0)\\
      \EE[YS|Z=1]&=\EE[\st]\mut0+\EE[\st^2](\mut1-\mut0)
    \end{split}
  \end{equation*}
\end{lemma}
Since $\st=\st^2$, the expression for $\EE[\st Y_T]$ in Lemma \ref{lemma:expectation} can be simplified to $\EE[\st]\mut1$; however, we retain the unsimplified version, since its similarity with the analogous expression when $Z_i=0$ simplifies the reasoning below.

@@

Now, the estimating equations $\sum_i\tilde{\Psi}_i=\bm{0}$ are algebraically equivalent to the estimating equations for a particular OLS fit, after some transformations, the most important of which follows.
Let
\begin{equation}\label{eq:ri}
\ri=Z_iS_i+(1-Z_i)\ppi=\begin{cases}
\ppi &Z_i=0\\
\sti &Z_i=1
\end{cases}
\end{equation}
Note that, in general, $(1-Z_i)\ppi=(1-Z_i)\ri$ and $Z_i\sti=Z_i\ri$, allowing $\ri$ to take the place of $\ppi$ and $S_i$ througout $\tilde{\Psi}_i$.

\begin{prop}\label{prop:reg1}
  Under Assumptions \ref{ass:sm}, \ref{ass:randomization}, \ref{ass:vps}, and \ref{ass:ci}, if %principal scores are known exactly,
  $\ri$ is defined as in \eqref{eq:ri} and the model
\begin{equation}\label{eq:regression0}
  \EE[Y_i|Z_i,\ri]=\beta_0+\beta_1\ri+\beta_2Z_i+\beta_3Z_i\ri
\end{equation}
is fit with OLS, yielding coefficient estimates $\hat{\beta}_k$, $k=0,\dots,3$, then
\begin{equation}
  \begin{split}
    \hat{\eff0}&\equiv \hat{\beta}_2\\
    \hat{\eff1}&\equiv \hat{\beta}_3-\hat{\beta}_2
  \end{split}
\end{equation}
are M-estimators, with $\hat{\eff0}\rightarrow_p\eff0$ and $\hat{\eff1}\rightarrow_p\eff1$ as $n\rightarrow\infty$
\end{prop}

@@

If principal scores were known exactly, the M-estimation structure of $\hat{\eff0}$ and $\hat{\eff1}$ implies that their sampling variance could be estimated by the standard sandwich estimator for OLS \citep{sefanskiBoosArticle}.
This will rarely be the case.
However, if the principal scores are modeled as \eqref{eq:prinScoreMod}, and parameters $\bm{\alpha}$ are estimated by M-estimation, with estimating equations $\sum_{i: Z_i=1}\Omega_i=\bm{0}$, then the estimating equations for the principal score model and principal effect estimation can be stacked \citep[c.f.][]{boosStefanskiBook}, as
\begin{equation}\label{eq:stacked}
  \sum_{i=1}^n \begin{pmatrix}
    Z_i\Omega_i(\bxsi,S_i;\bm{\alpha})\\
    \Psi_i\left(Y_i,Z_i,\ri(\bm{\alpha}); \bm{\beta}\right)\end{pmatrix}=\begin{pmatrix} \bm{0}\\\bm{0}\end{pmatrix}
\end{equation}
where $\ri\left(\bm{\alpha}\right)=S_iZ_i+\ppip{\bm{\alpha}}$ emphasizes $\ppi$'s dependence on $\bm{\alpha}$.
In practice, the principal score model \eqref{eq:prinScoreMod} could be fit first, yielding estimates $\bm{\hat{\alpha}}$, and model \eqref{eq:regression0} could be fit next, using $\ri=\ri\left(\bm{\hat{\alpha}}\right)$.
Together, the vector of estimates $[\bm{\hat{\alpha}},\bm{\hat{\beta}}]$ represents a zero of the stacked estimating equations \eqref{eq:stacked}.

@@

\subsection{Relaxing Covariate Ignorability with Outcome Modeling}
In most applications, identifying a set of covariates $\bxs$ satisfying CI will be difficult or impossible.
Fortunately, CI may be relaxed or obviated by regression.
Let $\bxy\in \mathbb{R}^k$, be a vector of covariates (possibly the same as $\bxs$); $\bxy$ can contain transformations and interactions of baseline covariates, allowing for a wide range of models.
Then note that $\st$ is, essentially, a covariate, so by randomization, (Assumption \ref{ass:rand}), $\EE[\bxy|\st]=\EE[\bxy|\st,Z=0]=\EE[\bxy|\st,Z=1]$.
Then information about the stratum-specific means of $\bxy$ can be extrapolated from the treatment to the control group.
This leads to the following proposition:
\begin{prop}\label{prop:interactions}
  Say
  \begin{equation}\label{eq:interaction}
    \begin{split}
      Y_{Ci}=\beta_0^C+\bm{\beta_1}^{C'}\bxy_i+\beta_2^C\sti+\bm{\beta_3^{C'}}\bxy_i\sti+\epsilon_{Ci}\\
      Y_{Ti}=\beta_0^T+\bm{\beta_1}^{T'}\bxy_i+\beta_2^T\sti+\bm{\beta_3^{T'}}\bxy_i\sti+\epsilon_{Ti}\\
    \end{split}
  %  \end{equation}
%      +\beta_3Z_i+\beta_4\sti Z_i+\bm{\beta_5}'\bxy_i\sti+
 %   \bm{\beta_6}'\bxy Z_i+\bm{\beta_7}'\bxy_i Z_i\sti+
  \end{equation}
  with $\EE[\epsilon_i]=0$, $\{Z_i,\sti,\bxy_i\}\independent \epsilon_i$.
  Say $\ppi$ linearly independent of $\bxy$, and
  that the following model is fit with OLS
  \begin{equation}\label{eq:fullInteractionModel}
    Y_i=\beta_0+\bm{\beta_1}'\bxy_i+\beta_2\ri+\beta_3Z_i+\beta_4\ri Z_i+\bm{\beta_5}'\bxy_i\ri+
    \bm{\beta_6}'\bxy Z_i+\bm{\beta_7}'\bxy_i Z_i\ri+\epsilon_i
  \end{equation}
  with $\ri$ as defined in \eqref{eq:ri},
  yielding coefficient estimates $\hat{\beta}_k$, $k=0,\dots,3$.
  Then
   \begin{equation}\label{eq:prinEffEst}
  \begin{split}
    \hat{\eff0}_{int}&\equiv \hat{\beta}_3+\bm{\beta}_6'\overline{\bxy}_{Z=1,S=0}\\
    \hat{\eff1}_{int}&\equiv \hat{\beta}_3+(\bm{\hat{\beta}_6}+\bm{\hat{\beta}_7})'\overline{\bxy}_{Z=1,S=1}
  \end{split}
\end{equation}
   are M-estimators, with $\hat{\eff0}_{int}\rightarrow_p\eff0$ and $\hat{\eff1}_{int}\rightarrow_p\eff1$ as $n\rightarrow\infty$,
   where $\overline{\bxy}_{Z=1,S=s}$ is a vector of the $k$ column means of the rows of $\bxy$ corresponding to subjects with $Z=1$ and $S=s$.
\end{prop}

Proposition \ref{prop:interactions} would seem to nearly solve the problem of estimating principal effects, or at least reducing it to the difficulty of other commen problems in causal modeling.
Unfortunately, simulation studies presented in the appendix suggest that, although consistent estimates for principal effects may be available, these estimates are far too inefficient for practical use; standard errors will be prohibitively high in all but the largest sample sizes.
For instance, if there is one standard normal covariate $x$, and $\pp_i=\left[1+exp\{-x_i/2\right]^{-1}$, then $Var\left(\hat{\muc0}_{int}\right)/Var(Y_C)\approx 8,500/n$.
In contrast, an analogous simulation in which Assumption \ref{ass:ci} holds, and the estimation routine of Proposition \ref{prop:reg1} was used, $Var\left(\hat{\muc0}_{int}\right)/Var(Y_C)\approx 15/n$.

Fortunately, simpler models can be fit much more efficiently. In particular, if interactions between covariates and treatment assignment are omitted, then data from the treatment group---in which $\st$ is fully observed---can be used to estimate the coefficients on covariates.
Such a model relies on what \citet{bo2002} calls ``Additivity of Treatment Assignment Effect,'' that $\bm{\beta_1}^{C}=\bm{\beta_1}^{T}$.
In other words, the only interactions between covariates and treatment assignment are due to differences between principal effects---a fairly strong assumpton.
However, there are two mitigating points.
First, since there is no requirement that all relevant covariates be included in the model, a researcher can exclude covariates from a model if he or she thinks their relationship with $Y$ may vary substantially between treatment groups, even in the same principal stratum.
Second, simulation results in the next section suggest that some models excluding important interactions may still yield reliable inferences.

>>>>

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
\usepackage{amsmath,amssymb}
\usepackage{bm}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{fullpage}
\usepackage{makecell}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{numdef}

@@

Randomized experiments easily admit subgroup analysis---estimation of treatment effects for a subset of the study population---when those subgroups are defined at baseline, and not themselves a function of treatment assignment.
A simple, popular approach to estimating subgroup effects is to use ordinary least squares (OLS) regression, regressing the outcome on a treatment indicator interacted with an indicator for subgroup membership.
However, when subgroup membership is a function of a post-treatment variable, then it may itself be affected by the treatment.
Contrasts between treatment and control conditions within those subgroups may not have a causal interpretation at all---and when they do, the same techniques that work for subgroups defined at baseline (such as OLS with interactions) will often produce inconsistent estimates.
Principal stratification \citep{frangakis} is a broad framework for defining subgroup effects for post-treatment subgroups, based on the notion of \emph{potential} subgroups called principal strata, that is, groups of subjects defined based on the values a post-treatment variable would potentially take if the subject is assigned to one treatment condition or another.
Since principal strata membership is based on subjects' responses to both factual and counter-factual treatment assignments, it is, in general unobserved.
Nevertheless, researchers have devised estimators of principal effects, as average effects within principal strata are called, to address a broad range of applied statistical problems, including non-compliance in randomized experiments, the evaluation of surrogate outcomes \citep{li2010bayesian}, the investigation of causal mechanisms \citep{lidsayPage}, and sample attrition \citep{zhangRubin, ding2011}.

Because principal strata membership is generally unobserved, principal effects are not always identified, and when they are, estimation can still be difficult.
Typically, principal effects are identified and estimated based on strong structural assumptions \citep[e.g.][]{air} and/or parametric modeling \citep[e.g.][]{imbens1997bayesian}.
Parametric models for principal effects are often hard to specify and hard or impossible to test.
Even when they are well-specified they can yield estimates that are severely biased and misleading \citep{griffin2008application,feller2016principal}.
Another route to identification and estimation of principal effects leads through auxiliary data, including baseline covariates or secondary outcomes \citep{mattei2013exploiting}.
Of particular importance is the ``principal score,'' \citep{jo,ding2017principal,feller2017principal}, or the probability that a subject belongs to a particular principal stratum, as a function of pre-treatment covariates.
Under a strong, untestable assumption called ``principal ignorability,'' which is akin to the no-confounding assumption underlying observational studies, principal scores can be transformed into weights, which in turn form an unbiased estimate of principal effects.

@@

This paper focuses on a special case of principal stratification---one-way, binary non-compliance---that is, there are only two principal strata, and principal stratum membership is observable in one treatment arm, but not in the other.
This scenario would occur if, say, subjects in the control arm of a study had no access to the treatment, and not all subjects assigned to the treatment arm actually received the full treatment.
While such a setup may represent the simplest and easiest principal stratification problem, we believe it is a common scenario.
Moreover, we believe that the methods developed here can extend to more complex scenarios, which we leave to future work.

@@

\section{Background}
In a randomized experiment with two conditions, let $Z_i=0$ if subject $i$, $i=1,\dots,n$ is assigned to the treatment condition, and let $Z_i=0$ otherwise.
If interest is in the effect of $Z$ on an outcome $Y$, under the ``stable unit treatment value assumption'' \citep{air} of no interference between units and no hidden versions of the treatment, define potential outcomes \citep{splawa1990application,air} $\yti$ as the value of $Y$ that $i$ would exhibit if $Z_i=1$, and let $\yci$ be the value if $Z_i=0$. Then the observed value of $Y$ satisfies $Y_i=Z_i\yti+(1-Z_i)\yci$, and the treatment effect of $Z$ on $Y$ for subject $i$ may be defined as $\tau_i=\yti-\yci$. The average treatment effect (ATE) is $\EE[\tau]$. %

Let $S_i\in\{0,1\}$ be a measurement on subject $i$ taken subsequent to treatment assignment, so that $S$ may be affected by $Z$.
Then $S$ itself has potential values $\sti$ and $\sci$, which $i$ would exibit if $Z_i=1$ or $Z_i=0$, respectively.
Assume ``strong monotonicity'' \citep[c.f.][]{dingLu} (also referred to as ``one-way noncompliance'' \citealt{}):
\begin{ass}[Strong Monotonicity]\label{ass:sm}
\begin{equation*}%\label{eq:strongMonotonicity}
  \sci=0\text{ for all }i
\end{equation*}
\end{ass}
Like potential outcomes $\yti$ and $\yci$, $\sci$ and $\sti$ are defined for all subjects, though $\sci$ is only observed when $Z_i=0$ and $\sti$ is only observed when $Z_i=1$. Under strong monotonicity, since $S$ is binary, every subject belongs to one of two ``principal strata,'' $\sti=1$ or $\sti=0$.
The average treatment effect in each principal stratum, $\eff1=\EE[\tau|\st=1]$ or $\eff1=\EE[\tau|\st=0]$, is called a principal effect. %
The challenge in estimating principal effects is due to the fact that strata membership $\sti$ is only observed when $Z_i=1$, in which case $\sti=S_i$, the observed value of $S$.

@@

Under Assumptions \ref{ass:sm} and \ref{ass:rand}, conditional means $\mut0$ and $\mut1$ are non-parametrically identified.
In particular, the means of observed outcomes for subjects with $Z=1$ and $S=0$ or $S=1$ are unbiased for $\mut0$ and $\mut1$, respectively, since when $Z_i=1$, $Y_i=\yti$ and $S_i=\sti$.
In contrast, $\muc0$ and $\muc1$ are not fully identified without further assumptions because $\sti$ and $\yci$ are never observed simultaneously (though they may be nonparametrically bounded; see, e.g., \citealt{bounding}).
When $S$ is a measure of compliance, and $\tau_i=0$ whenever $\sti=0$ (the ``exclusion restriction''), $\muc1$---and hence $\eff1$---may be estimated with instrumental variables methods \citep{air}.
In this case, $\eff1$ is often called the ``local average treatment effect'' or the average effect of the treatment on the treated.
We will briefly review two approaches to estimating $\muc0$ and $\muc1$ that do not assume the exclusion restriction---finite mixture modeling and weighting---with some comments on the role for covariates in-between.

@@

In order for principal scores to be useful, they need to vary with $\bm{x}$. Otherwise, $\bm{x}$ is uninformative about $\st$. In particular, we will assume the following:
\begin{ass}[Variable Principal Scores]\label{ass:vps}
 $\pp$ takes at least 3 distinct values
\end{ass}
In some estimators we will discuss below, two distinct values suffice. For a broader discussion, see \citet{ding2011,jiangDing2021}.

@@

Alternatively, assume ``principal ignorability'' \citep{jo2009use,dingLu}:
\begin{ass}[Principal Ignorability]\label{ass:PI}
\begin{equation}\label{eq:pi}
  Y_{C}\independent \st |\bxs
\end{equation}
\end{ass}
or a somewhat weaker version, $\EE[Y_C|\st,\bxs]=\EE[Y_C|\bxs]$ \citep{feller2017principal}.
That is, the principal stratum is unrelated to control potential outcomes conditional on covariates.
Principal ignorability \eqref{eq:pi} is reminiscent of ignorability assumptions typical in observational studies \citep[e.g.][]{rosenbaum2010design}.
Both require a rich set of covariates and background knowledge of the determinants of $S$; in particular, an unobserved covariate $u$ that is correlated with both $\st$ and $Y_C$ would invalidate \eqref{eq:pi}.

@@

We begin by introducing an assumption that we term ``Covariate Ignorability,'' or CI. It is a slighty weaker version of \citet{jiangDing2021}'s ``auxiliary independence'' assumption.
\begin{ass}[Covariate Ignorability]\label{ass:ci}
\begin{equation}\label{eq:assumption}
\EE[\yci|\bxsi,\sti]=\EE[\yci|\sti]=\muc0\text{ or }\muc1
\end{equation}
\end{ass}
i.e. $Y_C$ is mean-independent of $\bxs$ conditional on $\st$.
Under CI, covariates are not informative of the mean of $Y_C$ within principal strata.
As stated above, CI will rarely be plausible.
However, in some circumstances researchers can identify a subset of observed covariates that satisfy CI, and use those covariates in estimation.

It turns out that under strong monotonicity, randomization, and CI, and given a set of principal scores, the two principal effects $\eff0$ and $\eff1$ can be estimated by a simple OLS regression.
To see how, we'll %first
derive the estimating equations for $\muc0$, $\muc1$, $\mut0$, and $\mut1$, and show that they are equivalent to estimating equations for a model fit by OLS.% assuming that principal scores are known exactly, and then generalize to the case in which they are estimated using data from the treatment group.
The details of the calculations and the proofs are in the appendix.

The argument stems from a set of expressions for expectations, summarized in the following lemma:
\begin{lemma}\label{lemma:expectation}
  Under Assumptions \ref{ass:sm}, \ref{ass:rand}, \ref{ass:vps}, and \ref{ass:ci}, %if principal scores $\ppi$, for $i:Z_i=0$ are known exactly, then:
  \begin{equation*}
    \begin{split}
      \EE[Y|Z=0]=&\muc0+\EE\pp(\muc1-\muc0)\\
      \EE[Y\pp|Z=0]&=\EE[\pp]\muc0+\EE[\pp^2](\muc1-\muc0)\\
      \EE[Y|Z=1]&=\mut0+\EE[\st](\mut1-\mut0)\\
      \EE[YS|Z=1]&=\EE[\st]\mut0+\EE[\st^2](\mut1-\mut0)
    \end{split}
  \end{equation*}
\end{lemma}
Since $\st=\st^2$, the expression for $\EE[\st Y_T]$ in Lemma \ref{lemma:expectation} can be simplified to $\EE[\st]\mut1$; however, we retain the unsimplified version, since its similarity with the analogous expression when $Z_i=0$ simplifies the reasoning below.

@@

Now, the estimating equations $\sum_i\tilde{\Psi}_i=\bm{0}$ are algebraically equivalent to the estimating equations for a particular OLS fit, after some transformations, the most important of which follows.
Let
\begin{equation}\label{eq:ri}
\ri=Z_iS_i+(1-Z_i)\ppi=\begin{cases}
\ppi &Z_i=0\\
\sti &Z_i=1
\end{cases}
\end{equation}
Note that, in general, $(1-Z_i)\ppi=(1-Z_i)\ri$ and $Z_i\sti=Z_i\ri$, allowing $\ri$ to take the place of $\ppi$ and $S_i$ throughout $\tilde{\Psi}_i$.

\begin{prop}\label{prop:reg1}
  Under Assumptions \ref{ass:sm}, \ref{ass:rand}, \ref{ass:vps}, and \ref{ass:ci}, if %principal scores are known exactly,
  $\ri$ is defined as in \eqref{eq:ri} and the model
\begin{equation}\label{eq:regression0}
  \EE[Y_i|Z_i,\ri]=\beta_0+\beta_1\ri+\beta_2Z_i+\beta_3Z_i\ri
\end{equation}
is fit with OLS, yielding coefficient estimates $\hat{\beta}_k$, $k=0,\dots,3$, then
\begin{equation}
  \begin{split}
    \hat{\eff0}&\equiv \hat{\beta}_2\\
    \hat{\eff1}&\equiv \hat{\beta}_3-\hat{\beta}_2
  \end{split}
\end{equation}
are M-estimators, with $\hat{\eff0}\rightarrow_p\eff0$ and $\hat{\eff1}\rightarrow_p\eff1$ as $n\rightarrow\infty$
\end{prop}

@@

If principal scores were known exactly, the M-estimation structure of $\hat{\eff0}$ and $\hat{\eff1}$ implies that their sampling variance could be estimated by the standard sandwich estimator for OLS \citep{stefanskiBoos}.
This will rarely be the case.
However, if the principal scores are modeled as \eqref{eq:prinScoreMod}, and parameters $\bm{\alpha}$ are estimated by M-estimation, with estimating equations $\sum_{i: Z_i=1}\Omega_i=\bm{0}$, then the estimating equations for the principal score model and principal effect estimation can be stacked \citep[c.f.][]{boosStefanskiBook}, as
\begin{equation}\label{eq:stacked}
  \sum_{i=1}^n \begin{pmatrix}
    Z_i\Omega_i(\bxsi,S_i;\bm{\alpha})\\
    \Psi_i\left(Y_i,Z_i,\ri(\bm{\alpha}); \bm{\beta}\right)\end{pmatrix}=\begin{pmatrix} \bm{0}\\\bm{0}\end{pmatrix}
\end{equation}
where $\ri\left(\bm{\alpha}\right)=S_iZ_i+\ppip{\bm{\alpha}}(1-Z_i)$ emphasizes $\ppi$'s dependence on $\bm{\alpha}$.
In practice, the principal score model \eqref{eq:prinScoreMod} could be fit first, yielding estimates $\bm{\hat{\alpha}}$, and model \eqref{eq:regression0} could be fit next, using $\ri=\ri\left(\bm{\hat{\alpha}}\right)$.
Together, the vector of estimates $[\bm{\hat{\alpha}},\bm{\hat{\beta}}]$ represents a zero of the stacked estimating equations \eqref{eq:stacked}.

@@

\subsection{Relaxing Covariate Ignorability with Outcome Modeling}\label{sec:regression}
In most applications, identifying a set of covariates $\bxs$ satisfying CI will be difficult or impossible.
Fortunately, CI may be relaxed or obviated by regression (see \citealt[][\S 3.4]{jiangDing2021} for an analogous identification result).
Let $\bxy\in \mathbb{R}^k$, be a vector of covariates (possibly the same as $\bxs$); $\bxy$ can contain transformations and interactions of baseline covariates, allowing for a wide range of models.
Then note that $\st$ is, essentially, a covariate, so by randomization, (Assumption \ref{ass:rand}), $\EE[\bxy|\st]=\EE[\bxy|\st,Z=0]=\EE[\bxy|\st,Z=1]$.
Then information about the stratum-specific means of $\bxy$ can be extrapolated from the treatment to the control group.
This leads to the following proposition:
\begin{prop}\label{prop:interactions}
  Say, for $i=1,\dots,n$, principal scores $\ppi$ are generated as \eqref{eq:pscore}, with parameters $\bm{\alpha}$ identified and estimable with M-estimation, and outcomes are generated as
  \begin{equation}\label{eq:interaction}
    Y_i=\beta_0+\beta_1\sti+\beta_2 Z_i+\beta_3Z_i\sti+\bm{\gamma_1}'\bxy_i+\bm{\gamma_2}'\bxy_i\sti+
    \bm{\gamma_3}'\bxy Z_i+\bm{\gamma_4}'\bxy_i Z_i\sti+\epsilon_i
  \end{equation}
  with $\EE[\epsilon_i]=0$, $\{Z_i,\sti,\bxy_i\}\independent \epsilon_i$, and $\epsilon_i\independent\epsilon_j$ when $i\ne j$.

  Then, under Assumptions \ref{ass:sm}, \ref{ass:rand}, and \ref{ass:vps}, if $\ppi$ is linearly independent of $\bxy$, a researcher may follow the following procedure to estimate principal effects:
  \begin{enumerate}
  \item Estimate principal scores by fitting model \eqref{eq:pscore} to data from the treatment group
  \item Replace $\sti$ with $\ri$ (as defined in \ref{eq:ri}) in model \eqref{eq:interaction} and fit with OLS
  \item Estimate principal effects as:
   \begin{equation}\label{eq:prinEffEst}
  \begin{split}
    \hat{\eff0}_{int}&\equiv \hat{\beta}_3+\bm{\beta}_6'\overline{\bxy}_{Z=1,S=0}\\
    \hat{\eff1}_{int}&\equiv \hat{\beta}_3+(\bm{\hat{\beta}_6}+\bm{\hat{\beta}_7})'\overline{\bxy}_{Z=1,S=1}
  \end{split}
   \end{equation}
   where $\overline{\bxy}_{Z=1,S=0}$ and $\overline{\bxy}_{Z=1,S=0}$ are the vector of covariate sample means for the subsets of subjects with $Z=1$ and $S=0$ or $S=1$, respectively.
  \end{enumerate}
  Then $\hat{\eff0}_{G}$ and $\hat{\eff1}_{int}$ are M-estimators, with $\hat{\eff0}_{int}\rightarrow_p\eff0$ and $\hat{\eff1}_{G}\rightarrow_p\eff1$ as $n\rightarrow\infty$.
  Under suitible regularity conditions, they are jointly asymptotically normal, with a variance of the form \eqref{eq:sandwich}.
\end{prop}

In short, researchers can estimate principal effects under strong monotonicity by simply imputing missing $\st$ values with estimated principal scores, and fitting an OLS regression with appropriate interaction terms---just as if $\st$ were completely observed.

Proposition \ref{prop:interactions} would seem to nearly solve the problem of estimating principal effects under strong monotonicity, or at least seem to reduce it to the difficulty of other common problems in causal modeling.
Unfortunately, simulation studies presented in the appendix suggest that, although consistent estimates for principal effects may be available, these estimates are far too inefficient for practical use; standard errors will be prohibitively high in all but the largest sample sizes.
For instance, if there is one standard normal covariate $x$, and $\pp_i=\left[1+exp\{-x_i/2\right]^{-1}$, then $Var\left(\hat{\muc0}_{int}\right)/Var(Y_C)\approx 8,500/n$.
In contrast, an analogous simulation in which Assumption \ref{ass:ci} holds, and the estimation routine of Proposition \ref{prop:reg1} was used, $Var\left(\hat{\muc0}_{int}\right)/Var(Y_C)\approx 15/n$.
